Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-05 talk/dls2014/paper/paper.tex
introduction overhaul by trg

@@

\title{Virtual Memory Assisted Transactional Memory for Dynamic Languages}
\subtitle{DLS'14}

@@

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages
(e.g.\ PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g.\ Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads.

While this GIL prevents any parallelism from occurring, it also
provides some useful guarantees. Since this lock is always acquired
while executing bytecode instructions and it may only be released
in-between such instructions, it provides perfect isolation and
atomicity between multiple threads for a series of
instructions. Additionally, it provides the application with a
sequential consistency model~\cite{lamport79}. Another technology that
can provide the same guarantees is transactional memory (TM).

There have been several attempts at replacing the GIL with
TM~\cite{nicholas06,odaira14,fuad10}. Using transactions to enclose
multiple bytecode instructions, we can get the very same semantics as
the GIL while possibly executing several transactions in
parallel. Furthermore, by exposing these interpreter-level
transactions to the application in the form of \emph{atomic
blocks}~\cite{tim03,tim05}, we give dynamic languages a new
synchronisation mechanism that avoids several of the problems of locks
as they are used now.

TM systems can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In~\cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

\cfbolz{both the introduction and the contributions need to stress the co-design of GC and STM}

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well on low numbers of CPUs.
\item We make novel use of the memory management capabilities of
  off-the-shelf CPUs, e.g. memory segmentation and 64bit address space.
\item We integrate our STM system closely with a garbage collector
  (GC) in order to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

>>>>

\newcommand{\comment}[1]{}

@@

\title{
}
%\subtitle{DLS'14}

\comment{
A Platform for Parallelism in Dynamic languages
Parallelism for Python
A TM Implementation for Python
Transactional Memory as a foundation for parallel Python
Transactional Memory as a foundation for parallelism in Python
Parallel Python based on VM-assisted TM
Parallel Python based on VM-assisted Transactional Memory
A Platform for parallel Python
A TM Platform for parallel Python
A Transactional Memory Platform for parallel Python
A Transactional Memory system for parallel Python

Virtual Memory Assisted Transactional Memory for Dynamic Languages}

@@

Dynamic languages like Python, PHP, Ruby, and JavaScript receive a lot of attention but
have not yet embraced parallelism. A parallel programming model was not part of the design
of those languages. Thus, the reference implementations of, e.g., Python and Ruby use a
single, global interpreter lock (GIL) to serialise the execution of code in threads. The
use of a single global lock causes several disadvantages, and as multicore processors
become the de-facto standard in platfroms ranging from smart phones to datacenters, the
issue of parallel execution must be addressed by these dynamic languages.

Unfortunately the GIL plays a central role in current implementations. Even
recent just-in-time compilers (JIT) for these languages
(e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
are based on the GIL; Jython (an implementation of Python that is based on the JVM) is a
notable exception.

The standard model employed in these systems is fairly straightforward: there can be
multiple threads, but each acquires the GIL, then performs a single instruction, and
subsequently releases the lock.
While this setup limits the benefits due to parallelism
%% not clear we must say so TRG
% -- parallelism is limited to I/O or other activities external to the interpreter
the use of a GIL also provides some crucial guarantees. Since this lock is always acquired
while executing bytecode instructions and may be released only in-between such
instructions, the GIL ensures perfect isolation between multiple threads for a series of
instructions, which are executed atomically.  In addition, the GIL provides the
application with a sequential consistency model~\cite{lamport79}.

A global interpreter lock assures the atomic execution of individual instructions but
offers the programmer no help in managing concurrent data structures. Many parallel
programs must maintain invariants involving multiple objects (or even multiple fields of a
single object).  So to enable Python programs for parallel execution, we need a way to
identify which blocks of data must be executed atomically and hardware or software support
to effect this atomic execution.

\emph{Atomic
blocks}~\cite{tim03,tim05} allow dynamic languages to indicate
the units of computation that must be executed atomically to ensure correct concurrent
execution.  Atomic blocks map nicely to transactional memory, which provide a convenient
mechansim to replace the GIL.
 Using transactions to enclose
multiple bytecode instructions, we can get the very same semantics as
provided by the GIL while possibly executing several transactions in
parallel. And by exposing these interpreter-level
transactions to the application, we
avoid the well-known problems caused by locks as applications use atomic blocks instead
of explicit acquire/release operations on locks.

There have been several attempts at replacing the GIL with
TM~\cite{nicholas06,odaira14,fuad10}.
We focuse here on
software based transactional memory (STM)
because they provide the unique opportunity to consider the issues of atomicity,
synchronization, and memory management (resp. garbage collection) together.
STM systems impose no a-priori bound on the size of an atomic block and provide therefore
a flexibility that is absent from current hardware-based TM systems~\cite{wayforward14}.
The downside of STM systems is runtime overhead~\cite{cascaval08,drago11}, which can
be substantial despite recent efforts to lower the overhead~\cite{warmhoff13,spear09}.
In this paper, we describe how we manage to lower the overhead of an
STM system so that it can serve as a viable replacement for the GIL by leveraging the
demands of a dynamic language like Python to integrate the STM system with the garbage
collector.

\cfbolz{both the introduction and the contributions need to stress the co-design of GC and STM}
% agree trg

This paper makes the following contributions:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well for Python programs.
\item We integrate the  system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}
The STM described here is efficient and has low overhead and is therefore beneficial also
for environments with a small number of threads or cores. The paper describes also
%% can be dropped, not the core topic
how to introduce atomic blocks into the Python language to provide a
  backwards compatible, composable synchronization mechanism for
  threads.

%% What is really needed from this section?? Maybe the subsection on Python can be moved
%%  to the intro -- the rest is questionable.  TRG
